Medical Imaging Classification with Cold-Atom Reservoir Computing using Auto-Encoders and Surrogate-Driven Training\\

### 2025-05-27 paper/main.tex
draft

\begin{abstract}
We introduce a quantum-classical hybrid 
architecture for medical image classification 
based on neutral atom quantum processors. 
This approach is designed to address the challenges of 
medical imaging, with a particular focus on tasks such 
as polyp detection and classification.
By integrating an autoencoder guided by a quantum 
reservoir, the pipeline learns compact and discriminative 
representations of image data that are also well-suited for quantum reservoir 
computing. To overcome the non-differentiability of quantum 
measurements, we circumvent this `gradient barrier' 
by incorporating a differentiable surrogate model that simulates the 
behaviour of the quantum layer, enabling end-to-end backpropagation. 
The guided training process jointly optimizes for both image 
reconstruction and classification accuracy, ensuring that the latent 
representations are both meaningful and effective for quantum processing. 
In our implementation, image data is encoded as atom detuning parameters 
in a Rydberg Hamiltonian, and quantum embeddings are obtained through expectation 
values. These embeddings are then passed to a linear classifier, 
enabling faster training and inference compared to deep classical networks.
Our experiments show that this method outperforms traditional 
approaches using PCA or unguided autoencoders. We also conduct 
ablation studies to evaluate the impact of quantum and training 
parameters, demonstrating the robustness and flexibility of the 
proposed pipeline for real-world medical imaging applications, 
even in the NISQ era.
\end{abstract}

@@

%============================================
% INTRODUCTION
%============================================
\section{Introduction}

@@

Quantum computing offers novel opportunities 
for machine learning, particularly through quantum reservoir 
computing (QRC), where a physical quantum system processes 
classical inputs into high-dimensional nonlinear embeddings~\cite{tanakaRecentAdvancesPhysical2019,fujiiHarnessingDisorderedQuantum2017}. Recent works show that analog quantum systems, 
such as neutral-atom platforms, can serve as untrained 
reservoirs with rich dynamics for temporal and pattern 
recognition tasks \cite{domingoOptimalQuantumReservoir2022,kornjavcaLargescaleQuantumReservoir2024}. 
In hybrid approaches, a classical encoder compresses 
image data, and a quantum reservoir expands 
the encoded features into a higher-dimensional space, 
potentially boosting classification performance.

A major challenge in such hybrid quantum-classical 
models is the non-differentiability of quantum measurements, 
which obstructs gradient-based optimization. Additionally, 
tuning quantum parameters can suffer from barren plateaus, where 
gradients vanish in high-dimensional Hilbert spaces \cite{mccleanBarrenPlateausQuantum2018}. 
To address this, we introduce a classical neural surrogate 
that emulates the quantum reservoir's input-output behavior. 
This surrogate enables end-to-end training via backpropagation, 
while the quantum system remains fixed and non-trainable.

@@

\input{reservoir_architecture.tex}

\subsection{Quantum Reservoir Computing}
Quantum Reservoir Computing (QRC) extends the reservoir 
computing paradigm into the quantum domain. By leveraging quantum systems' 
inherent properties, such as superposition and entanglement, QRC aims 
to enhance computational capabilities. Implementations using quantum 
oscillators have shown promise in solving complex learning tasks, 
offering advantages over classical counterparts. Notably, large-scale 
experiments utilizing neutral-atom analog quantum computers have demonstrated 
the scalability and effectiveness of QRC in various machine learning applications~\cite{kornjavcaLargescaleQuantumReservoir2024}

@@

\subsection{Reservoir Computing with Neutral Atoms}
As Quantum Reservoir Computing (QRC) leverages the complex 
dynamics of quantum systems to process information, 
extending the Classical Reservoir Computing 
paradigm into the quantum domain. Neutral atom platforms, 
particularly those utilizing Rydberg states, 
have emerged as promising candidates for implementing 
QRC due to their scalability and controllable interactions.

In the work by M. Kornjača et al.~\cite{kornjavcaLargescaleQuantumReservoir2024}, 
a large-scale, gradient-free QRC algorithm was 
developed and experimentally implemented on a 
neutral-atom analog quantum computer. 
This system achieved competitive performance across 
various machine learning tasks, 
including classification and time-series prediction, 
demonstrating effective learning with increasing system 
sizes up to 108 qubits.

@@

\input{autoencoder_architecture.tex}

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{System Architecture Overview}

Our proposed pipeline integrates classical and quantum 
components for dimensionality reduction and classification. 
A classical autoencoder compresses high-dimensional image 
data into a lower-dimensional latent space. This latent 
representation is encoded into a Rydberg atom chain, 
serving as input to a quantum reservoir. The reservoir's 
dynamics, governed by the Rydberg Hamiltonian, produce 
quantum embeddings through measurements of quantum observables.
The overall architecture is illustrated in `Fig.-\ref{fig:qgars_pipeline}'.

To facilitate end-to-end training, we employ a differentiable 
surrogate model that approximates the quantum reservoir's 
behavior, enabling gradient flow from the classification 
loss back to the autoencoder. The surrogate's outputs feed 
into a linear classifier trained to map quantum embeddings to class labels.

The autoencoder is trained to minimize a combined loss 
function that interpolates between reconstruction loss and 
classification loss, promoting representations that are both 
informative for input reconstruction and effective for quantum 
reservoir computing.

Once trained, the encoder compresses data into the latent 
space, which is then processed by the actual quantum reservoir. 
The resulting quantum embeddings are used to train a linear 
classifier for final predictions.

\begin{figure*}[!h]
    \centering
    \resizebox{0.9\textwidth}{!}{
        \includegraphics{qgars_pipeline_cropped.pdf}
    }
    \label{fig:qgars_pipeline}
    \caption{Overview of the Quantum-Guided Autoencoder Reservoir Computing System.}
\end{figure*}

\subsection{Quantum Guided Autoencoder}
The Quantum Guided Autoencoder (QGA) combines classical 
preprocessing with quantum processing to leverage 
the strengths of both paradigms. The classical encoder 
reduces the dimensionality of the input data, facilitating 
efficient quantum processing. The quantum circuit 
then processes the compressed data, aiming to reconstruct 
the original input and perform classification simultaneously.

@@

\begin{equation}
    \mathcal{L} = (1-\lambda) \cdot \mathcal{L}_R + \lambda \cdot \mathcal{L}_{C},
    \label{eq:composed_loss_function}
\end{equation}
where:
\begin{itemize}
    \item \( \mathcal{L}_R = \| x - \hat{x} \|^2 \) is the reconstruction loss, measuring the mean squared error between the input \( x \) and its reconstruction \( \hat{x} \).
    \item \( \mathcal{L}_C = -\sum_i y_i \log(\hat{y}_i) \) is the classification loss, computed as the cross-entropy between the true labels \( y_i \) and the predicted probabilities \( \hat{y}_i \).
    \item \( 0 < \lambda < 1 \) is a hyperparameter that controls the trade-off between reconstruction and classification objectives.
\end{itemize}

%\subsubsection{Balancing Reconstruction and Classification}

\subsection{The Gradient Barrier Problem}
Hybrid quantum-classical models, such as the Quantum Guided Autoencoder 
(QGAE), aim to leverage quantum computational advantages within 
classical machine learning frameworks. However, a significant 
challenge arises due to the non-differentiable nature of quantum 
operations, commonly referred to as the `gradient barrier'.

@@

\subsubsection{Architecture}
In our pipeline, the surrogate model is implemented as a 
feedforward neural network designed to learn the mapping 
from the autoencoder's latent space to the quantum 
embeddings produced by the quantum reservoir.

Its architecture comprises multiple fully connected 
layers with nonlinear activation functions, facilitating 
the approximation of the complex, non-linear 
transformations inherent in quantum dynamics.

@@

\subsubsection{Training Procedure}
\begin{itemize}
    \item Data Input: 
    Feed a batch of latent data \( z \) from the autoencoder 
    into the real quantum reservoir, recording the resulting quantum 
    embeddings.

    \item Surrogate Training:
    Train the surrogate model using the recorded quantum
    embeddings as targets, minimizing the loss function:
    \begin{equation}
        \mathcal{L}_{\text{surrogate}} = \| \hat{y} - y \|^2,
    \end{equation}
    where \( \hat{y} \) are the surrogate model's predictions
    and \( y \) are the actual quantum embeddings.

    \item Regular Updates:
    Periodically update the surrogate model during training
    to insure it accurately reflects the quantum reservoir's
    behavior. This can be done after a fixed number of epochs
    or when the surrogate's performance on a validation set
    reaches a certain threshold.
\end{itemize}

@@

As a results, we not only achieve end-to-end training of the
hybrid model but also makes it computationally efficient,
as we reduce the need for frequent quantum measurements.

@@

\subsubsection{Quantum Readout Methods}
The quantum reservoir's output is obtained by probing, 
in successive time steps, the state of the atoms 
after the systems evolution up to a certain time \( t \).

The readout is performed by measuring specific observables,
such as $\langle Z_i \rangle$, $\langle Z_i \, Z_j\rangle$, and $\langle Z_i \, Z_j \, Z_k \rangle$, which measure respectively one-, two- and three-qubit correlations.
% \begin{itemize}
%     \item \textbf{Single-atom Measurements}: 
%     \( \langle Z_i \rangle \) for each atom \( i \), where \( Z_i \) is the Pauli-Z operator.
%     \item \textbf{Two-atom Correlations}: 
%     \( \langle Z_i Z_j \rangle \) for pairs of atoms \( (i, j) \), capturing correlations between their states.
%     \item \textbf{Three-atom Correlations}:
%     \( \langle Z_i Z_j Z_k \rangle \) for triplets of atoms \( (i, j, k) \), providing higher-order correlations.
% \end{itemize}

@@

\subsubsection{Quantum Reservoir Computing Layer}
Our quantum reservoir computing implementation uses 
a Rydberg atom simulator with the following parameterizable
components:

@@

\subsubsection{Classical Network Architectures}
After the traning process of the quantum-guided autoencoder,
the following quantum embeddings are passed to a linear classifier.
The classifier is a simple feedforward neural network with the following architecture:
\begin{itemize}
    \item A single hidden layer with a ReLU activation function.
    \item Cross-entropy loss for training, optimized using the Adam optimizer.
\end{itemize}

\subsection{Comparison Methods}
We benchmarked our model against some
classical methods, including:
\begin{itemize}
    \item \textbf{PCA + Linear Classifier:} 
    A linear classifier trained on the PCA-reduced features.

    \item \textbf{Autoencoder + Linear Classifier:} 
    A linear classifier trained on the latent representations from the autoencoder.

    \item \textbf{PCA + Neural Network:}
    A fully connected neural network trained on the PCA-reduced features.

    \item \textbf{Autoencoder + Neural Network:}
    A fully connected neural network trained on the latent representations from the autoencoder.
\end{itemize}

@@

\subsection{Parameter Sweep Strategy}
To optimize the performance of our quantum-guided autoencoder, 
we conducted a systematic parameter sweep over the following hyperparameters:
\begin{itemize}
    \item \textbf{Guided Lambda Parameter (\( \lambda \))}: 
    This parameter controls the trade-off between reconstruction and classification objectives in the loss function~\ref{eq:composed_loss_function}. We tested values ranging from 0.1 to 0.9 in increments of 0.1.

@@

    \item \textbf{Quantum Parameters:} 
    Including the number of atoms in the chain, Rabi frequency, and detuning parameters. We varied these parameters to assess their impact on classification performance.
    \item \textbf{Batch Size:}
    \item \textbf{Learning Rate:}
\end{itemize}

%============================================
% RESULTS
%============================================
\section{Results and Discussion}
\subsection{Classification Performance Comparison}
\subsection{Ablation Studies}
\subsubsection{Impact of Guided Lambda Parameter}
\subsubsection{Effect of Quantum Update Frequency}
\subsubsection{Influence of Quantum Parameters}
\subsection{Dimensionality Reduction Comparison}
\subsection{Surrogate Model Fidelity Analysis}
\subsection{Generalization to Unseen Data}

%============================================
% LIMITATIONS AND FUTURE WORK
%============================================
\section{Limitations and Future Work}
\subsection{Current Limitations}
\subsection{Potential Extensions}
\subsection{Hardware Implementation Considerations}

@@

@@

\newpage
\newpage
\tableofcontents

>>>>

\begin{abstract}
We introduce a quantum-classical hybrid
architecture for medical image classification based on neutral-atom quantum processors. 
This approach is designed to address the challenges of
medical imaging, with a particular focus on tasks such
as polyp detection and classification.
By integrating an autoencoder guided by a quantum 
reservoir, the pipeline learns compact and discriminative 
representations of image data that are also well-suited for quantum reservoir 
computing. To overcome the non-differentiability of quantum 
measurements, we circumvent this `gradient barrier' 
by incorporating a differentiable surrogate model that simulates the 
behaviour of the quantum layer, enabling end-to-end backpropagation. 
The guided training process jointly optimizes for both image 
reconstruction and classification accuracy, ensuring that the latent 
representations are both meaningful and effective for quantum processing. 
In our implementation, image data is encoded as atom detuning parameters 
in a Rydberg Hamiltonian, and quantum embeddings are obtained through expectation 
values. These embeddings are then passed to a linear classifier, 
enabling faster training and inference compared to deep classical networks.
Our experiments show that this method outperforms traditional 
approaches using PCA or unguided autoencoders. We also conduct 
ablation studies to evaluate the impact of quantum and training 
parameters, demonstrating the robustness and flexibility of the 
proposed pipeline for real-world medical imaging applications, 
even in the NISQ era.
\end{abstract}

@@

%============================================
% hilbert
%============================================
\section{Introduction}

@@

Quantum computing offers novel opportunities 
for machine learning, particularly through quantum reservoir 
computing (QRC), where a physical quantum system processes 
classical inputs into high-dimensional nonlinear embeddings~\cite{tanakaRecentAdvancesPhysical2019,fujiiHarnessingDisorderedQuantum2017}. Recent works show that analog quantum systems, such as neutral-atom platforms, can serve as untrained reservoirs with rich dynamics for temporal and pattern recognition tasks~\cite{domingoOptimalQuantumReservoir2022,kornjavcaLargescaleQuantumReservoir2024}.

In hybrid approaches, a classical encoder compresses image data, and a quantum reservoir expands the encoded features into a higher-dimensional Hilbert space, potentially boosting classification performance, as these expanded features might capture high-order correlations and nonlinearities that classical methods cannot tractably represent.

A major challenge in such hybrid quantum-classical models is the non-differentiability of quantum measurements, which obstructs gradient-based optimization. Additionally, tuning quantum parameters can suffer from barren plateaus, where gradients vanish in high-dimensional Hilbert spaces~\cite{mccleanBarrenPlateausQuantum2018}. To address this, we introduce a classical neural surrogate that emulates the quantum reservoir's input-output behavior. 
This surrogate enables end-to-end training via backpropagation, while the quantum system remains fixed and non-trainable and is only used to train the surrogate model and in downstream classifiers.

@@

\input{images/diagrams/reservoir_architecture.tex}

\subsection{Quantum Reservoir Computing}

\subsubsection{Main Differences}
Quantum Reservoir Computing (QRC) extends the reservoir 
computing paradigm into the quantum domain. By leveraging quantum systems' 
inherent properties, such as superposition and entanglement, QRC aims 
to enhance computational capabilities. Implementations using quantum 
oscillators have shown promise in solving complex learning tasks, 
offering advantages over classical counterparts. Notably, large-scale 
experiments utilizing neutral-atom analog quantum computers have demonstrated 
the scalability and effectiveness of QRC in various machine learning applications~\cite{kornjavcaLargescaleQuantumReservoir2024}

@@

\subsubsection{Quantum Reservoir Computing with Neutral Atoms}
Neutral atom platforms, particularly those utilizing Rydberg states, have emerged as promising candidates for implementing QRC due to their scalability and controllable interactions.

In the work by M. Kornjača et al.~\cite{kornjavcaLargescaleQuantumReservoir2024}, a large-scale, gradient-free QRC algorithm was developed and experimentally implemented on a neutral-atom analog quantum computer. This system achieved competitive performance across various machine learning tasks, including classification and time-series prediction, demonstrating effective learning with increasing system sizes up to $108$ qubits.

@@

\input{images/diagrams/autoencoder_architecture.tex}

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{System Architecture Overview}
Our proposed pipeline is a Quantum-Guided Autoencoder with a 
Reservoir Surrogate (QGARS). It is a hybrid architecture that 
synergizes classical autoencoding techniques with quantum 
reservoir processing to achieve efficient learning of features
that work well with the reservoir computing layer whose embeddings
will be used to perferm a classification task.
The system is designed to overcome the inherent non-differentiability of quantum operations, which traditionally impede gradient-based training methods.
The overall architecture is illustrated in `Fig.-\ref{fig:qgars_pipeline}'.

In a higher level, the \textbf{architectural components} are the following:
\begin{itemize}
    \item \textbf{Classical Autoencoder:}
        \begin{itemize}
            \item \textbf{Encoder:}
            The encoder network \(\varepsilon_\omega(x)\)
            transforms high-dimensional input data \(\mathbf{x} \in \mathbb{R}^{n}\) into 
            a lower-dimensional latent representation \(\mathbf{z} \in \mathbb{R}^m\), 
            where \(m < n\).
            \item \textbf{Decoder:}
            The decoder network \(\mathcal{D}_\rho(z)\)
            Reconstructs the input data from the latent representation, producing
            \(\mathbf{\hat{x}} \in \mathbb{R}\).
        \end{itemize}

    \item \textbf{Quantum Reservoir Layer:}
        \begin{itemize}
            \item \textbf{Parameter Mapping:}
            \(g(\mathbf{z})\) maps the latent representation \(\mathbf{z}\) to the local detuning frequencies
            \(\Delta_{\mathrm{i}}\) of the Rydberg Atoms.
            \item \textbf{Quantum Evolution:}
            Evolves the system under a Hamiltonian \(H(\Delta)\), resulting in a quantum 
            state \(|\psi(\Delta)\rangle\).
            \item \textbf{Measurement:}
            Performs measurements on the quantum state to obtain embeddings \(\mathbf{q} \in \mathbb{R}^k\), which serve as training targets for the Surrogate Model or as 
            inputs for the classification task later on.
        \end{itemize}

    \item \textbf{Surrogate Model:}
    A differentiable classical neural network \( f_{\text{sur}}(\mathbf{z}; \theta_{\text{sur}}) \)
    where \(\theta_{\text{sur}}\), which are it's parameters, are tuned to approximate the 
    mapping from \(\mathbf{z}\) to \(\mathbf{q}\), effectively emulating
    the quantum layer's behaviour.

    \item \textbf{Linear Classifier:}
    A classical neural network that maps the quantum embeddings \(\mathbf{q}\) to 
    predicted class labels \(\hat{y}\).

\end{itemize}
The following sections will give a more detailed explanation of each 
component.

\begin{figure*}[!ht]
    \centering
    \resizebox{0.9\textwidth}{!}{
        \includegraphics{images/diagrams/qgars_pipeline_cropped.pdf}
    }
    \label{fig:qgars_pipeline}
    \caption{Overview of the Quantum-Guided Autoencoder Reservoir Computing System.}
\end{figure*}

\subsection{Quantum Guided Autoencoder}
The Quantum Guided Autoencoder (QGA) combines classical 
preprocessing with quantum processing to leverage 
the strengths of both paradigms.

Ideally, the latent space representation created by the encoder \(\varepsilon_\omega(x)\)
would be passed through two sections of the model:
\begin{itemize}
    \item \textbf{The Decoder Network \(\mathcal{D}_\rho(z)\):}
    that tries to reconstruct the original input data.
    \item \textbf{The Quantum Reservoir:}
    that extracts highly dimensional quantum embeddings and
    passes them to a linear classifier to perform a classification
    task and map the embeddings to the image labels.
\end{itemize}
However, some problems emerge when using a real quantum reservoir directly during the autoencoder training stage, which will be discussed in a later section.

@@

\begin{equation}
    \mathcal{L} = (1-\lambda) \cdot \mathcal{L}_R + \lambda \cdot \mathcal{L}_{C},
    \label{eq:composed_loss_function}
\end{equation}
where:
\begin{itemize}
    \item \( \mathcal{L}_R = \| \mathbf{x} - \hat{\mathbf{x}} \|^2 \) is the reconstruction loss, measuring the mean squared error between the input \( \mathbf{x} \) and its reconstruction \( \mathbf{\hat{x}} \).
    \item \( \mathcal{L}_C = -\sum_i y_i \log(\hat{y}_i) \) is the classification loss, computed as the cross-entropy between the true labels \( y_i \) and the predicted probabilities \( \hat{y}_i \), which are obtained through a linear mapping of the 
    approximated embeddings \(\mathbf{q}'\).
    \item \( 0 < \lambda < 1 \) is a hyperparameter that controls the trade-off between reconstruction and classification objectives.
\end{itemize}    
\subsection{The Gradient Barrier Problem}
Hybrid quantum-classical models, such as the Quantum Guided Autoencoder 
(QGAE), aim to leverage quantum computational advantages within 
classical machine learning frameworks. However, a significant g
challenge arises due to the non-differentiable nature of quantum 
operations, commonly referred to as the `gradient barrier'.

@@

\subsubsection{Surrogate Model Architecture}
In our pipeline, the surrogate model \( f_{\text{sur}}(\mathbf{z}; \theta_{\text{sur}}) \) 
is implemented as a feedforward neural network 
designed to learn the mapping 
from the autoencoder's latent space to the quantum 
embeddings produced by the quantum reservoir. 
\begin{equation}
    f_{\text{sur}}(\mathbf{z}; \theta_{\text{sur}}) = \mathbf{q}' \approx \mathbf{q} = \mathrm{Measure}(|\psi(g(\mathbf{z})\rangle)
\end{equation}
Its architecture comprises multiple fully connected 
layers with nonlinear activation functions, facilitating 
the approximation of the complex, non-linear 
transformations inherent in quantum dynamics.

@@

\subsubsection{Training Procedure}
\begin{itemize}
    \item Data Input: 
    Feed a batch of latent data \( \mathbf{z} \) from the autoencoder 
    into the real quantum reservoir, recording the resulting quantum 
    embeddings.

    \item Surrogate Training:
    Train the surrogate model  by minimizing the loss function:
    \begin{equation}
        \mathcal{L}_{\text{sur}} = \| q' - q \|^2,
    \end{equation}

    \item Regular Updates:
    Periodically update the surrogate model during training
    to ensure it accurately reflects the quantum reservoir's
    behavior. This can be done after a fixed number of epochs
    or when the surrogate's performance on a validation set
    reaches a certain threshold.
\end{itemize}

@@

As a result, we not only achieve end-to-end training of the
hybrid model but also makes it computationally efficient,
as we reduce the need for frequent quantum measurements.

@@

\subsubsection{Quantum Readout Methods}
The quantum reservoir's output is obtained by probing, 
in successive time steps, the state of the atoms 
after the systems evolution up to a certain time \( t \).
The readout is performed by measuring specific observables,
such as $\langle Z_i \rangle$, $\langle Z_i \, Z_j\rangle$, and $\langle Z_i \, Z_j \, Z_k \rangle$, which measure respectively one-, two- and three-qubit correlations.
% \begin{itemize}
%     \item \textbf{Single-atom Measurements}: 
%     \( \langle Z_i \rangle \) for each atom \( i \), where \( Z_i \) is the Pauli-Z operator.
%     \item \textbf{Two-atom Correlations}: 
%     \( \langle Z_i Z_j \rangle \) for pairs of atoms \( (i, j) \), capturing correlations between their states.
%     \item \textbf{Three-atom Correlations}:
%     \( \langle Z_i Z_j Z_k \rangle \) for triplets of atoms \( (i, j, k) \), providing higher-order correlations.
% \end{itemize}

@@

\subsection{Quantum Reservoir Computing Layer}
Our quantum reservoir computing implementation uses 
a Rydberg atom simulator with the following parameterizable
components:

@@

\subsection{QRC Embeddings Linear Classifier}
After the traning process of the QGARS model,
the following quantum embeddings are passed to a linear classifier.
The classifier is a simple feedforward neural network with the following architecture:
\begin{itemize}
    \item A single hidden layer with a ReLU activation function.
    \item Cross-entropy loss for training, optimized using the Adam optimizer.
\end{itemize}

\subsection{Comparison Methods}
We benchmarked our model against some
classical methods, including:
\begin{itemize}
    \item \textbf{PCA/Autoencoder + Linear Classifier:} 
    A linear classifier trained on the PCA-reduced features or the latent representations
    from the autoencoder.

    \item \textbf{PCA/Autoencoder + Neural Network:}
    A fully connected neural network trained on the features from the 
    same reduction methods as the linear classifier. 
\end{itemize}

We also benchmarked our QGARS encoding strategy
against the performance of the Linear Mapping of
QRC embeddings extracted from the latent representation 
of data using other feature reduction methods, such as
PCA and the Classical Autoencoder.

@@

\subsection{Parameter Sweep Strategy}
To optimize the performance of our quantum-guided autoencoder, 
we conducted a systematic parameter sweep over the following hyperparameters:
\begin{itemize}
    \item \textbf{Guided Lambda Parameter (\( \lambda \))}: 
    To analyse the effect that the trade-off between reconstruction and classification objectives in the loss function~\ref{eq:composed_loss_function} affects
    the QRC

@@

    \item \textbf{Quantum Parameters:} 
    Rabi frequency, and detuning parameters. We varied these parameters to assess their impact on classification performance.

    \item \textbf{Number of Features:}
    The number of features is the same as the number of qubits (atoms), 
    as each feature encoded in each atom.

\end{itemize}
The results from these experiments will be discussed in the next section.

%============================================
% RESULTS
%============================================
\section{Results and Discussion}

\subsection{Classification Performance Comparison}
We first compare classification accuracy of the Linear Classifier on the QRC embeddings across PCA,
vanilla autoencoder, and our QGARS pipeline on each dataset. 
Table~\ref{tab:qrc_reduction_accuracy} summarizes mean accuracy and standard deviation over 
five runs, using 12 features.

\begin{table}[!htb]
  \caption{Classification accuracy (\%) on test sets}
  \label{tab:qrc_reduction_accuracy}
  \centering
  \begin{tabular}{lccc}
    \hline
    Method & Synthetic Polyp & CVC-ClinicDB & MNIST(0/1) \\
    \hline
    PCA + QRC + Linear      &  00 ± 0.0  &  00 ± 0.0  &  00 ± 0.0 \\
    AE + QRC + Linear       &  00 ± 0.0  &  00 ± 0.0  &  00 ± 0.0 \\
    QGARS + QRC + Linear    &  \textbf{00 ± 0.0} &  \textbf{00 ± 0.0} &  \textbf{00 ± 0.0} \\
    \hline
  \end{tabular}
\end{table}

\subsection{Dimensionality Reduction Comparison}

To evaluate how well the quantum reservoir embeddings separate classes, we project the high-dimensional embeddings down to $2$ dimensions via t-distributed Stochastic Neighbour Embedding (t-SNE). Figure~\ref{fig:embeddings_tsne} shows side-by-side plots for PCA → QRC, AE → QRC, and QGARS → QRC pipelines. The QGARS embeddings exhibit the tightest and most distinct clusters, indicating superior discriminative structure.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.32\textwidth]{pca_qrc_tsne.pdf}
  \includegraphics[width=0.32\textwidth]{ae_qrc_tsne.pdf}
  \includegraphics[width=0.32\textwidth]{qgars_qrc_tsne.pdf}
  \caption{t-SNE projections of the 4 700-D QRC embeddings for (left) PCA + QRC, (center) AE + QRC, (right) QGARS + QRC.}
  \label{fig:embeddings_tsne}
\end{figure*}

Quantitatively, we compute the Silhouette Score on the original 4 700-D embeddings (Table~\ref{tab:silhouette}). The QGARS pipeline achieves the highest score, confirming better inter-class separation:

\begin{table}[ht]
  \caption{Silhouette Scores for QRC embeddings}
  \label{tab:silhouette}
  \centering
  \begin{tabular}{lcc}
    \hline
    Pipeline         & Silhouette Score \\
    \hline
    PCA + QRC        & 0.0 \\
    AE + QRC         & 0.0 \\
    QGARS + QRC      & \textbf{0.0} \\
    \hline
  \end{tabular}
\end{table}

\subsection{Ablation Studies}

\subsubsection{Impact of Guided Lambda Parameter}
We vary the guidance weighting $\lambda$ in Eq.~\ref{eq:composed_loss_function} to assess its influence on performance. Figure~\ref{fig:lambda_sweep} plots test accuracy vs.\ $\lambda$.

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{paper/images/results/generated_polyp_dataset/guided_lambda_log_effect.pdf}
  \caption{Accuracy as a function of the reconstruction/classification trade-off parameter $\lambda$.}
  \label{fig:lambda_sweep}
\end{figure}

\noindent
We find optimal performance around $\lambda=XXX$; too little guidance ($\lambda<XXX$) yields under-trained classifiers, while too much ($\lambda>XXX$) degrades reconstruction quality.

\subsubsection{Effect of Quantum Update Frequency}

\subsubsection{Influence of Quantum Parameters}

\subsection{Surrogate Model Fidelity Analysis}
We assess how well the surrogate approximates the quantum reservoir by 
computing the MSE between true quantum embeddings and surrogate outputs over the test set (Figure~\ref{fig:surrogate_fidelity}). Low MSE indicates high fidelity.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=0.8\linewidth]{surrogate_mse.pdf}
  \caption{Surrogate vs.\ quantum embedding MSE over training epochs.}
  \label{fig:surrogate_fidelity}
\end{figure}

\subsection{Comparison with Classical Methods}
To benchmark the benefit of our QGARS pipeline, 
we compare against purely classical feature-reduction 
approaches followed by standard classifiers. 
Table~\ref{tab:classical_comp} reports test accuracy on each dataset 
for the different methods.

\begin{table}[ht]
  \caption{Test set classification accuracy (\%) for classical baselines}
  \label{tab:classical_comp}
  \centering
  \begin{tabular}{lccc}
    \hline
    Method            & Synthetic Polyp & CVC-ClinicDB & MNIST (0/1) \\
    \hline
    PCA + Linear      &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    AE + Linear       &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    AE + NN           &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    \hline
  \end{tabular}
\end{table}

\noindent
In contrast, the QGARS pipeline achieves 0.0\%, 0.0\%, and 0.0\% on the same datasets (see Table~\ref{tab:qrc_reduction_accuracy}).

\subsubsection{Training Loss Dynamics}

Figure~\ref{fig:loss_curves} shows the training loss curves for the classical baselines and QGARS on the Synthetic Polyp dataset. We plot the reconstruction loss for PCA/AE methods combined with classification loss for the AE+NN and QGARS pipelines.

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{loss_comparison.pdf}
  \caption{Normalized training loss vs.\ epoch for PCA+Linear, AE+Linear, AE+NN, and QGARS on the Synthetic Polyp dataset.}
  \label{fig:loss_curves}
\end{figure}

\noindent
Although the loss function of the QGARS method shows
signs of instability, is not only converges faster but also reaches a 
lower combined loss

%============================================
% LIMITATIONS AND FUTURE WORK
%============================================
\section{Limitations and Future Work}
\subsection{Current Limitations}
While QGARS demonstrates strong performance, several limitations remain:
\begin{itemize}
    \item \textbf{Surrogate Fidelity:}
    Imperfect surrogate approximation can introduce bias in the gradients and limit end‐to‐end 
    performance, especially if the latent distribution shifts significantly during training.
    \item \textbf{Classical overhead:}  
    Frequent surrogate retraining and batch quantum simulations introduce 
    nontrivial classical compute overhead, potentially offsetting some of the 
    quantum advantage.
\end{itemize}

\subsection{Potential Extensions}
Several avenues can further enhance QGARS:
\begin{itemize}
    \item \textbf{Different Hamiltonians:}
    Explore alternative interaction Hamiltonians (e.g., XY, Heisenberg) to better tailor reservoir dynamics, as experimented by A. D. Lorenzis et al~\cite{lorenzisHarnessingQuantumExtreme2025}.

    \item \textbf{Convolutional Autoencoders:}
    Replace the fully connected autoencoder with a convolutional AE to learn spatially localized features, making the model input‐size agnostic and enabling richer local structure for the reservoir.

    \item \textbf{Reservoir encoding schemes:}
    Investigate alternate encodings—such as position encoding via atom placement, global versus local pulse shaping, or multi‐tone driving—to diversify the reservoir’s nonlinear response.

    \item \textbf{\(\lambda\)-Scheduling \& Curriculum Learning:}  
    Implement a curriculum that gradually increases the guidance weight \(\lambda\) from 0 to 1, allowing the autoencoder to first stabilize reconstruction before introducing quantum guidance. Hybrid ramping schedules may accelerate convergence.

    \item \textbf{Noise‐Aware Training:}  
    Incorporate realistic noise models (decoherence, measurement error) into surrogate training so that the surrogate learns hardware‐like quantum embeddings, improving robustness upon deployment in NISQ devices.

\end{itemize}

\subsection{Hardware Implementation Considerations}
Bringing QGARS to physical hardware entails additional challenges:
\begin{itemize}
    \item \textbf{Calibration and drift:}  
        Rydberg atom platforms require careful calibration of laser frequencies and intensities; drift over time may degrade reservoir fidelity.
    \item \textbf{Shot noise and sampling cost:}  
        Finite‐shot measurements introduce statistical noise; optimizing shot budgets versus surrogate fidelity is critical.
    \item \textbf{Scalability:}  
        Extending to hundreds of atoms demands sophisticated trap control and cross‐talk mitigation; resource scheduling on shared QPUs may limit throughput.
    \item \textbf{Integration with classical pipelines:}  
        Efficient data transfer between classical and quantum subsystems (e.g.\ via FPGA or GPU–QPU interconnects) and asynchronous surrogate updates will be essential for low‐latency operation.
\end{itemize}

@@

We have presented QGARS, a hybrid quantum-classical architecture that integrates a classical autoencoder with a neutral-atom quantum reservoir and a differentiable surrogate model to enable end‐to‐end training. By jointly optimizing reconstruction and classification losses, QGARS learns compact and discriminative features tailored for quantum reservoir computing. Our experiments on synthetic polyp, CVC-ClinicDB, and binary MNIST datasets demonstrate that QGARS outperforms classical PCA‐ and autoencoder‐based baselines. Ablation studies highlight the importance of the guidance weight \(\lambda\), surrogate update frequency, and reservoir parameters. While challenges remain—particularly regarding high‐dimensional embeddings, surrogate fidelity, and classical overhead, our proposed extensions and hardware considerations chart a clear path toward robust, scalable quantum‐enhanced medical imaging pipelines in the NISQ era.

@@
